---
title: "Estimating panel conditioning with a refreshment sample"
author: "Shoki Okubo"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Estimating panel conditioning with a refreshment sample}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A panel survey that adds a refreshment sample can compare respondents who have
been interviewed $k$ times (the continuing cohort) with first-time respondents
drawn from the same population (the fresh cohort) at the same wave. The
difference is the panel conditioning effect of dose $k$ among the continuing
cohort's survivors, provided the fresh cohort is a valid control. This vignette
runs the designs of the package on simulated data. The notation and the
identification results are those of Okubo (2026), "Panel conditioning as a
dose-response causal effect: identification with refreshment samples".

## Data layout

`pc_estimate()` takes two data frames. `old` has one row per member of the
continuing cohort with the wave-$t$ answer `y` (`NA` for non-respondents), the
cohort's own entry-wave answer `y_entry`, the survival indicator `s_prior`
(responded at every wave from entry to $t$), optionally `s_next` (responds at
$t+1$) and a discrete covariate `x`. `new` has one row per first-wave respondent
of the fresh cohort with `y`, `s_m` (responds at the $m = k$ waves after $t$)
and optionally `s_m1` (responds at $k + 1$ further waves). `pc_from_wide()`
builds both from a wide file with per-wave response indicators.

`pc_simulate()` generates both cohorts from the data-generating process of the
paper's Monte Carlo study. Under the `MNAR_trait` regime, dropout depends on an
unobserved persistent trait that also enters the outcome, so the survivors are a
selected group and the naive contrast is biased.

```{r}
library(panelcond)
set.seed(1)
sim <- pc_simulate(n_old = 4800, n_new = 960, k = 4, regime = "MNAR_trait")
sim$truth
```

`pce_surv` is the estimand of the naive, survival-matched and entry-corrected
designs (the effect among the continuing cohort's survivors); `pce_surv_next` is
the estimand of the symmetric design; `pce_pop` is the population effect that
IPW targets.

## Estimates and diagnostics

```{r}
fit <- pc_estimate(sim$old, sim$new, k = 4, adjust = "x", nboot = 200)
fit
```

The naive contrast is biased downward by the survivors' selection differential
(reported as `entry-wave selection differential`); survival matching (`sm`),
symmetric survival matching (`ssm`), the entry-wave correction (`ec`) and its
regression-standardised version (`ec_adj`) recover the survivor effect. The
`se` column holds the first-order analytic standard errors; `se_boot` the joint
person bootstrap, in which persons are resampled within each cohort and every
estimator is recomputed, so that the diagnostic differences $T_1$ (SM against
EC) and $T_2$ (SSM against SM) receive bootstrap denominators as well. In this
regime attrition is stationary and depends on the trait, not on the current
state, so SM, SSM and EC share the same estimand up to a small composition term
and both tests have their nominal size. $T_1$ does not reject; $T_2$ falls at
the 5% boundary in this particular draw, which is a rejection at the nominal
error rate, not evidence of state dependence (the Monte Carlo section below
shows the rejection rates under a regime where state dependence is present).

## Sensitivity

`pc_bounds()` gives worst-case bounds on the population effect, obtained by
assigning the non-survivors the endpoints of the outcome's support, and the
breakdown values of the two main designs: the SM estimate in fresh-cohort
standard-deviation units, and the additive drift in the selection differential
at which the EC estimate vanishes or its 95% interval reaches zero.

```{r}
pc_bounds(fit, y_range = c(-4, 4), sd_new = sd(sim$new$y))
```

## Two continuing cohorts

When a later refreshment sample targets a different population it cannot serve
as a dose-0 control, but two continuing cohorts drawn from the same population
at different entry waves identify the increment between their doses, each
corrected by its own entry wave.

```{r}
set.seed(2)
a <- pc_simulate(n_old = 4800, n_new = 10, k = 12, regime = "MNAR_trait")
b <- pc_simulate(n_old = 960,  n_new = 10, k = 8,  regime = "MNAR_trait")
inc <- pc_increment(data.frame(y = a$old$y, y_entry = a$old$y_entry, s = a$old$s_prior),
                    data.frame(y = b$old$y, y_entry = b$old$y_entry, s = b$old$s_prior),
                    kA = 12, kB = 8, nboot = 200)
inc
a$truth[["pce_surv"]] - b$truth[["pce_surv"]]
```

## Design planning

`pc_montecarlo()` repeats the simulation and reports, for each estimator, the
bias against its own estimand, the Monte Carlo standard deviation, the mean
analytic standard error and the coverage of the nominal 95% interval; the
rejection rates of the two diagnostic tests are attached as an attribute.

```{r}
set.seed(3)
mc <- pc_montecarlo(R = 100, n_old = 4800, n_new = 960, k = 4, regime = "MNAR_state")
print(mc, digits = 3)
attr(mc, "tests")
```

Under state-dependent attrition, survival matching and the entry-wave
correction are biased by the same amount and $T_1$ cannot see it, whereas the
symmetric design is unbiased and $T_2$ rejects.
